---
title: "R-Ladies RTP ggplot2 Workshop"
author: "Elaine McVey"
date: "October 15, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome to R-Ladies RTP!

## Outline

- Why ggplot2?
- Conceptual ggplot2
- Basic ggplot2 syntax
- Common pitfalls
- The power of layers
- The power of groups
- The power of facets
- Scales and legends
- Themes and refinements
- Extensions

## Materials

- All source code at https://github.com/rladies/rtp-data-viz-ggplot2-10-15-16

- Slides at 

## Datasets

- Datasets for today are here: https://www.dropbox.com/s/zhmn02ti0ggxdj7/rladies_ggplot2_datasets.rda?dl=1

  - You can download them from R:
  
      `download.file(
      'https://www.dropbox.com/s/zhmn02ti0ggxdj7/
       rladies_ggplot2_datasets.rda?dl=1',
              'rladies_ggplot2_datasets.rda')`
              
      `attach('rladies_ggplot2_datasets.rda')`
  
## Dataset Descriptions

- Transit ride data
  - rides_df: one row per ride
  - daily_df: daily summary of rides
  - hourly_df: hourly summary of rides
  
- Durham registered voter data
  - durham_voters_df: one row per voter

# Why ggplot2?

## Exploratory Power

- explore data very quickly, and don't screw up
- graphics "pit of success"
- refine plots to be publication quality  

## Conceptually Cohesive

- based on the grammar of graphics
- principles and patterns
- very complex plots from simple building blocks

## Dominance

- the most prominent static graphics package in R
- help is easy to find
- developers are contibuting extensions

# Conceptual ggplot2



# Basic ggplot2 syntax

## The required elements

- DATA
- MAPPING
- GEOM

## Our first hands-on example ...

```{r, message=FALSE} 
download.file(
'https://www.dropbox.com/s/zhmn02ti0ggxdj7/rladies_ggplot2_datasets.rda?dl=1',
        'rladies_ggplot2_datasets.rda')
        
attach('rladies_ggplot2_datasets.rda')

library(tidyverse)
```


## Basic plot


```{r}

ggplot(data = daily_df) +
geom_point(mapping = aes(x = ride_date, y = n_rides))

```

## Mappings can be general

```{r}

ggplot(data = daily_df, mapping = aes(x = ride_date, y = n_rides)) +
geom_point()

```

## Parameters can be unnamed

```{r}

ggplot(daily_df, aes(x = ride_date, y = n_rides)) +
geom_point()

```

## Data can be passed in

```{r, fig.height = 4}

daily_df %>%
ggplot(aes(x = ride_date, y = n_rides)) +
geom_point()

```

## Data can be passed in

```{r, fig.height = 4}

daily_df %>%
ggplot(aes(x = ride_date, y = n_rides)) +
geom_point()

```

## Mapping is more than x and y

```{r, fig.height = 4}

daily_df %>%
ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
geom_point()

```

## Mapping is more than x and y

```{r, fig.height = 4}

daily_df %>%
ggplot(aes(x = ride_date, y = n_rides, size = n_riders)) +
geom_point()

```

## Variable creation on the fly...

```{r, fig.height = 4}

daily_df %>%
ggplot(aes(x = ride_date, y = n_rides, color = day_of_week %in% c('Sat', 'Sun'))) +
geom_point()

```

## ... or passed in

```{r, fig.height = 3}

daily_df %>%
  mutate(day_type = if_else(day_of_week %in% c('Sat', 'Sun'),
                            'Weekend',
                            'Weekday')) %>%
ggplot(aes(x = ride_date, y = n_rides, color = day_type)) +
geom_point()

```

# Common early pitfalls

## Mappings that aren't

```{r, fig.height = 4}

daily_df %>%
ggplot() +
geom_point(aes(x = ride_date, y = n_rides, color = 'blue'))

```

## Mappings that aren't

```{r, fig.height = 4}

daily_df %>%
ggplot() +
geom_point(aes(x = ride_date, y = n_rides), color = 'blue')

```

## + and %>%

`
daily_df %>%
  mutate(day_type = if_else(day_of_week %in% c('Sat', 'Sun'),
                            'Weekend',
                            'Weekday')) %>%
ggplot(aes(x = ride_date, y = n_rides, color = day_type)) %>%
geom_point()
`

# The power of layers

# The power of groups 

# The power of facets

# Everyday conveniences

## Labels and titles

# Scales and legends

# Themes and refinements

# Extensions