---
title: "R-Ladies RTP ggplot2 Workshop"
author: "Elaine McVey"
date: "October 15, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome to R-Ladies RTP!

## Outline

- Why ggplot2?
- Conceptual ggplot2
- Basic ggplot2 syntax
- Common pitfalls
- The power of layers
- The power of groups
- The power of facets
- Scales and legends
- Themes and refinements
- Extensions

## Materials

- All source code at https://github.com/rladies/rtp-data-viz-ggplot2-10-15-16

## Datasets

- Datasets for today are here: https://www.dropbox.com/s/zhmn02ti0ggxdj7/rladies_ggplot2_datasets.rda?dl=1

  - You can download them from R:
  
      `download.file(
      'https://www.dropbox.com/s/zhmn02ti0ggxdj7/
       rladies_ggplot2_datasets.rda?dl=1',
              'rladies_ggplot2_datasets.rda')`
              
      `attach('rladies_ggplot2_datasets.rda')`
  
## Dataset Descriptions

- Transit ride data
  - rides_df: one row per ride
  - daily_df: daily summary of rides
  - hourly_df: hourly summary of rides
  
- Durham registered voter data
  - durham_voters_df: one row per voter

# Why ggplot2?

## Exploratory Power

- explore data very quickly, and don't screw up
- graphics "pit of success"
- refine plots to be publication quality  

## Conceptually Cohesive

- based on the grammar of graphics
- principles and patterns
- very complex plots from simple building blocks

## Dominance

- the most prominent static graphics package in R
- help is easy to find
- developers are contibuting extensions

# Conceptual ggplot2

## The basics

- map variables to aethestics
- add "geoms" for visual representation layers
- scales can be independently managed
- legends are automatically created
- statistics are sometimes calculated by geoms...

## Learn more

- ggplot2 book 

# Basic ggplot2 syntax

## The required elements

- DATA
- MAPPING
- GEOM

## Our first hands-on example ...

```{r, message=FALSE} 
download.file(
'https://www.dropbox.com/s/zhmn02ti0ggxdj7/rladies_ggplot2_datasets.rda?dl=1',
        'rladies_ggplot2_datasets.rda')
        
attach('rladies_ggplot2_datasets.rda')

library(tidyverse)
```


## Basic plot


```{r}

ggplot(data = daily_df) +
geom_point(mapping = aes(x = ride_date, y = n_rides))

```

## Mappings can be general

```{r}

ggplot(data = daily_df, mapping = aes(x = ride_date, y = n_rides)) +
geom_point()

```

## Parameters can be unnamed

```{r}

ggplot(daily_df, aes(x = ride_date, y = n_rides)) +
geom_point()

```

## Data can be passed in

```{r, fig.height = 4}

daily_df %>%
ggplot(aes(x = ride_date, y = n_rides)) +
geom_point()

```

## Data can be passed in

```{r, fig.height = 4}

daily_df %>%
ggplot(aes(x = ride_date, y = n_rides)) +
geom_point()

```

## Mapping is more than x and y

```{r, fig.height = 4}

daily_df %>%
ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
geom_point()

```

## Mapping is more than x and y

```{r, fig.height = 4}

daily_df %>%
ggplot(aes(x = ride_date, y = n_rides, size = n_riders)) +
geom_point()

```

## Variable creation on the fly...

```{r, fig.height = 4}

daily_df %>%
ggplot(aes(x = ride_date, y = n_rides, color = day_of_week %in% c('Sat', 'Sun'))) +
geom_point()

```

## ... or passed in

```{r, fig.height = 3}

daily_df %>%
  mutate(day_type = if_else(day_of_week %in% c('Sat', 'Sun'),
                            'Weekend',
                            'Weekday')) %>%
ggplot(aes(x = ride_date, y = n_rides, color = day_type)) +
geom_point()

```

# Common early pitfalls

## Mappings that aren't

```{r, fig.height = 4}

daily_df %>%
ggplot() +
geom_point(aes(x = ride_date, y = n_rides, color = 'blue'))

```

## Mappings that aren't

```{r, fig.height = 4}

daily_df %>%
ggplot() +
geom_point(aes(x = ride_date, y = n_rides), color = 'blue')

```

## + and %>%

`daily_df %>%`
  `mutate(day_type = if_else(day_of_week %in% c('Sat', 'Sun'),
                            'Weekend',
                            'Weekday')) %>%`
`ggplot(aes(x = ride_date, y = n_rides, color = day_type)) %>%`
`geom_point()`

## Basic plot exercise

Plot the number of unique routes per day over time, colored by day of week. (n_unique_routes)

## Basic plot exercise

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_unique_routes, color = day_of_week)) +
  geom_point() 
```



# The power of layers 

## Basic plot

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides)) +
  geom_point() 
```

## Two layers!
```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides)) +
  geom_point() +
  geom_line()
```

## Iterate on layers
```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides)) +
  geom_point() + 
  geom_smooth(span = .1) # try changing span
```

## The power of groups
```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() + 
  geom_line()
```

## Now we've got it
```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_smooth(span = .2, se = FALSE)
```

## Control data by layer
```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point(data = filter(daily_df, !(day_of_week %in% c('Sat', 'Sun')) & n_rides < 200),
             size = 5, color = 'gray') +
  geom_point()
```

## Control data by layer {.smaller}
```{r, fig.height = 3}

low_weekdays_df <- daily_df %>%
  filter(!(day_of_week %in% c('Sat', 'Sun')) & n_rides < 100)

daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, 
             color = day_of_week, label = ride_date)) +
  geom_point(data = low_weekdays_df, size = 5, color = 'gray') +
  geom_text(data = low_weekdays_df, aes(y = n_rides + 15), 
            size = 2, color = 'black') +
  geom_point() 
```

# The power of facets

## facet_wrap
```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides)) +
  geom_point() + 
  geom_line() +
  facet_wrap( ~ day_of_week)
```

## facet_grid {.smaller}
```{r, fig.height = 3}
durham_voters_df %>%
  group_by(race_code, gender_code, age) %>%
  summarize(n_voters = n(),
            n_rep = sum(party == 'REP')) %>%
  filter(gender_code %in% c('F','M') & 
           race_code %in% c('W', 'B', 'A') &
           age != 'Age < 18 Or Invalid Birth Date') %>%
  ggplot(aes(x = age, y = n_voters)) +
  geom_bar(stat = 'identity') +
  facet_grid(race_code ~ gender_code)
```

## Free scales {.smaller}
```{r, fig.height = 3}
durham_voters_df %>%
  group_by(race_code, gender_code, age) %>%
  summarize(n_voters = n(),
            n_rep = sum(party == 'REP')) %>%
  filter(gender_code %in% c('F','M') & 
           race_code %in% c('W', 'B', 'A') &
           age != 'Age < 18 Or Invalid Birth Date') %>%
  ggplot(aes(x = age, y = n_voters)) +
  geom_bar(stat = 'identity') +
  facet_grid(race_code ~ gender_code, scales = 'free_y')
```

## Facets + layers {.smaller}
```{r, fig.height = 4, echo = FALSE}
durham_voters_df %>%
  group_by(race_code, gender_code, age) %>%
  summarize(n_voters = n(),
            n_rep = sum(party == 'REP')) %>%
  filter(gender_code %in% c('F','M') & 
           race_code %in% c('W', 'B', 'A') &
           age != 'Age < 18 Or Invalid Birth Date') %>%
  mutate(age_cat = as.numeric(as.factor(age))) %>%
  ggplot(aes(x = age, y = n_voters)) +
  geom_point() +
  geom_line(aes(x = age_cat)) +
  geom_line(aes(x = age_cat, y = n_rep), color = 'red') +
  geom_point(aes(y = n_rep), color = 'red') +
  facet_grid(race_code ~ gender_code, scales = 'free_y') +
  expand_limits(y = 0) 
```

## Facets + layers {.smaller}

Note: better to use gather

```{r, fig.height = 3, eval = FALSE}
durham_voters_df %>%
  group_by(race_code, gender_code, age) %>%
  summarize(n_voters = n(),
            n_rep = sum(party == 'REP')) %>%
  filter(gender_code %in% c('F','M') & 
           race_code %in% c('W', 'B', 'A') &
           age != 'Age < 18 Or Invalid Birth Date') %>%
  mutate(age_cat = as.numeric(as.factor(age))) %>%
  ggplot(aes(x = age, y = n_voters)) +
  geom_point() +
  geom_line(aes(x = age_cat)) +
  geom_line(aes(x = age_cat, y = n_rep), color = 'red') +
  geom_point(aes(y = n_rep), color = 'red') +
  facet_grid(race_code ~ gender_code, scales = 'free_y') +
  expand_limits(y = 0) 
```



# Everyday conveniences

## Labels and titles

```{r, fig.height = 3}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_smooth(span = .2, se = FALSE) +
  xlab('') +
  ylab('# of Transit Rides') +
  ggtitle('Transit Rides over time by Day of Week') +
  scale_color_discrete('Day of Week')
```

# Scales and legends

## Scale transformation
```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  scale_y_reverse()
```

## Scale transformation

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  scale_y_sqrt()
```

## Scale details

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  scale_y_continuous(breaks = c(0, 200, 500))
```

# Themes and refinements

## Overall Themes

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  theme_bw()
```

## Overall Themes

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  theme_dark()
```

## Changing Details

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90))
```

## Themes Vignette

To really master themes:

- http://docs.ggplot2.org/dev/vignettes/themes.html

# Documentation

## Online Docs

Better than the regular documentation:

- http://docs.ggplot2.org/current/

# Extensions

## Contributed extensions

Check it out!

- http://www.ggplot2-exts.org/

# Questions?

## Next Meetup

Monday, November 14th, 6pm

Location:

- TransLoc, near RTP
- 4505 Emperor Blvd, Durham
- Parking is easy!

Topic:

- Getting data _into_ R

Topic/format suggestions?